---
title: "main_notebook"
author: "Maxim Musikhin"
date: "30 November 2020"
output: html_document
---

# Import Packages

```{r Packages Used, include=FALSE}
library(tidyverse)
library(rhdf5)
library(reshape2)
```

# From HDF5 to CSV
The original raw data I had was stored in HDF5 format.

```{r}
# The HDF5 files are named like this D1-PRO-R1001-CG-2lcb-110_129-AA.h5 The difference is only the 2 letters that identify the amino acids mutated.

#Assign the input and output directories
#input_directory <- "data//raw_data/"
#output_directory <- "data//extracted_data/"


convert_and_save_hdf5_as_csv <-
  function(input_directory, output_directory) {
    fileNames <- list.files(input_directory)
    
    for (fileName in fileNames) {
      aminoAcid <- substr(fileName, 30, 31)
      testHDF5 <-
        H5Fopen(paste(input_directory, fileName, sep = ""))
      df <- data.frame(t(testHDF5$d1))
      names(df) <- c("time", "distance")
      
      df$simulation <- case_when(df$time <= (100000 * 5) ~ 1,
                                 df$time <= (200000 * 5) ~ 2,
                                 df$time <= (299998 * 5) ~ 3)
      write.csv(
        x = df,
        file = paste(output_directory, paste(aminoAcid, ".csv", sep = ""), sep = ""),
        row.names = FALSE
      )
    }
  }



#convert_and_save_hdf5_as_csv(input_directory, output_directory)
```

# Setting up for work

## Initialising constants

```{r}
# Identify the path to the directory that contains the extracted data
data_directory <- "data//extracted_data/"
```

## Sort out the corrupted files

All the corrupted files should be filtered out.

This include files that:

- have more than 299,998 rows
- contain distance values that are out of allowed range (0.6-2.0) 

```{r}
identify_valid_files <- function(path_to_files) {
  # State criteria that will be used to select valid records
  allowed_records_number <- 299998
  allowed_distance_lower_limit <- 0.6
  allowed_distance_upper_limit <- 2.0
  
  # List the csv file names
  files <- list.files(path_to_files)
  
  # Initialise an array for correct files
  valid_files <- c()
  
  
  # Iterate over each file and check if they satisfy the condition
  for (i in seq_along(files)) {
    df <- read.csv(paste(path_to_files, files[i], sep = ""))
    
    df_cropped <- subset(
      df,
      df$distance > allowed_distance_lower_limit &
        df$distance < allowed_distance_upper_limit
    )
    
    number_of_valid_records <- nrow(df_cropped)
    
    if (number_of_valid_records == allowed_records_number) {
      valid_files <- append(valid_files, files[i])
    }
  }
  return(valid_files)
}
```

## Creating the main dataframe

Create data frame that will include the metadata for all of mutants.
```{r}
fill_the_main_df <- function() {
  # Get file names
  files <- identify_valid_files(data_directory)
  
  # Remove ".csv" ending to get only amino acid names
  aa_names <- substr(files, 1, 2) 
  
  # vectorise by letters
  fl <- substr(aa_names, 1, 1)
  sl <- substr(aa_names, 2, 2)
  
  # Create a data frame that will be filled later
  main_df <-
    data.frame(name = aa_names,
               first_letter = fl,
               second_letter = sl)
  
  # Fill it with amino acid property details
  # Hydrophobicity scale data (http://assets.geneious.com/manual/8.0/GeneiousManualsu41.html)
  hydrophobicity_df <- read.csv("data//hydrophobicity.csv")
  names(hydrophobicity_df) <- c("amino_acid", "hydrophobicity")
  # Other properties of amino acids (https://en.wikipedia.org/wiki/Proteinogenic_amino_acid)
  properties_df <- read.csv("data//chemical_props.csv")
  
  #to get the total "score" of a property I decided to add the values of two individual amino acid for each mutant
  for (i in 1:nrow(main_df)) {
    # Hydrophobicity score is a value that defines the relative hydrophobicity or hydrophilicity of amino acid residue
    main_df$hydrophobicity[i] <-
      hydrophobicity_df[hydrophobicity_df$amino_acid == main_df$first_letter[i], 2] +
      hydrophobicity_df[hydrophobicity_df$amino_acid == main_df$second_letter[i], 2]
    # The isoelectronic point or isoionic point is the pH at which the amino acid does not migrate in an electric field (it is the pH at which the amino acid is neutral)
    main_df$pi[i] <-
      as.numeric(as.character(properties_df[properties_df$Short == main_df$first_letter[i], 5])) +
      as.numeric(as.character(properties_df[properties_df$Short == main_df$second_letter[i], 5]))
    # Masses of amino acids
    main_df$mass[i] <-
      as.numeric(as.character(properties_df[properties_df$Short == main_df$first_letter[i], 4])) +
      as.numeric(as.character(properties_df[properties_df$Short == main_df$second_letter[i], 4]))
    #Van der Waals volume 
    main_df$vdw_volume[i] <-
      as.numeric(as.character(properties_df[properties_df$Short == main_df$first_letter[i], 8])) +
      as.numeric(as.character(properties_df[properties_df$Short == main_df$second_letter[i], 8]))
  }
  
  return(main_df)
}
```

```{r}
# Apply the function
main_df <- fill_the_main_df()
```

## Helper functions to easily get the data given the amino acids and simulation number

```{r}
# The function that reads an individual simulation of a given mutant
get_sim <- function(AAname, simNum, data_directory = "data//extracted_data/") {
  if (AAname %in% main_df$name) {
    fileName <- paste(AAname, ".csv", sep = "")
    path <- paste(data_directory, fileName, sep = "")
    
    df <- read.csv(path)
    
    if (simNum == 1) {
      return(df[c(10001:100000), ])
    } else if (simNum == 2) {
      return(df[c(110001:200000), ])
    } else if (simNum == 3) {
      return(df[c(209999:299998), ])
    }
  }
  else {
    return(0)
  }
}

# Function that gives all simulations concatenated
get_all_sim <- function(AAname, data_directory = "data//extracted_data/") {
  if (AAname %in% main_df$name) {
    fileName <- paste(AAname, ".csv", sep = "")
    path <- paste(data_directory, fileName, sep = "")
    df <- read.csv(path)
    
    
    return(rbind(df[c(c(10001:100000), c(110001:200000), c(209999:299998)), ]))
  }
  else {
    return(0)
  }
}
```

# Generic Plotting functions 

```{r}
# the function that gives a distance density plot of a sim
  distance_density_plot <- function(df, aminoAcids) {
  
  df$simulation <- as.factor(df$simulation)
  p <- ggplot(data = df, aes(x = distance, color = simulation)) +
    ggtitle(aminoAcids) +
    geom_density() + 
    scale_x_continuous(name="Distance", limits=c(0.6, 2.0), breaks=seq(0.6,2.0,0.2)) + 
    theme_minimal()
  return(p)
}
```

```{r}
# An example of use
distance_density_plot(get_all_sim("DC"), "DC") 
```
# Prediction Linear regression model

```{r}
#Separate each simulation of each mutant for easier analysis
sims_df <- rbind(data.frame(name = main_df$name, simulation = rep(1, length(main_df$name))),data.frame(name = main_df$name, simulation = rep(2, length(main_df$name))),data.frame(name = main_df$name, simulation = rep(3, length(main_df$name))))
sims_df <- sims_df[order(sims_df$name, sims_df$simulation),]
```

```{r}
#calculate the variance for each simulation

for (i in seq_along(sims_df$name)) {
  sim_data <- get_sim(sims_df$name[i],sims_df$simulation[i])
  sims_df$variance[i] <- var(sim_data$distance)
}


```

```{r}
for (i in seq_along(main_df$name)) {
  current_name <- main_df$name[i]
  main_df$total_var[i] <- var(get_all_sim(current_name)$distance)
}
```


```{r}
#mean variation and not the variation of whole mutant because due to unknown reasons (MD simulation related) they can start and get stuck at different distances, therefore it will not be representative
for (i in seq_along(main_df$name)) {
  current_name <- main_df$name[i]
  main_df$mean_var[i] <- mean(sims_df[sims_df$name == current_name,]$variance)
  main_df$var_of_var[i] <- var(sims_df[sims_df$name == current_name,]$variance)
  
}
```

```{r}
#generalisation for each amino acid mutation and its position
aa_df <- data.frame(amino_acid = unique(main_df$first_letter))

for (i in seq_along(aa_df$amino_acid)) {
  aa_df$as_first_var[i] <- mean(main_df[main_df$first_letter == aa_df$amino_acid[i],]$mean_var)
  aa_df$as_second_var[i] <- mean(main_df[main_df$second_letter == aa_df$amino_acid[i],]$mean_var)
  
  #aa_df$as_first_var[i] <- mean(main_df[main_df$first_letter == aa_df$amino_acid[i],]$var_of_var)
  #aa_df$as_second_var[i] <- mean(main_df[main_df$second_letter == aa_df$amino_acid[i],]$var_of_var)
  #
  #aa_df$as_first_var[i] <- mean(main_df[main_df$first_letter == aa_df$amino_acid[i],]$total_var)
  #aa_df$as_second_var[i] <- mean(main_df[main_df$second_letter == aa_df$amino_acid[i],]$total_var)
}

for (i in seq_along(main_df$name)) {
main_df$var_score[i] <-
      (aa_df[aa_df$amino_acid == main_df$first_letter[i],]$as_first_var +
      aa_df[aa_df$amino_acid == main_df$second_letter[i],]$as_second_var)
}

```

#Fluctuation analysis

```{r}
fp <- "data//bnz_distance/"  # specify file path
fl <- list.files( path = fp, pattern = "*.out", full.names = TRUE ) # get .out files
for (file in fl ){  # loop through files, read it and write it as .csv files. Then delete .out files
flag <- tryCatch( {
    write.table( x = read.table(file = file, header = FALSE ), 
                 file = file.path( fp, gsub( "out$", "csv", basename( file ) ) ), 
                 sep = ",", 
                 row.names = FALSE )
    TRUE
    },
    error = function( x ) { 
      print( paste0( "Problem reading and writing file : ", file ) )
      return(FALSE) } )

#if( flag ) unlink( file )  # deletes .out files
}
```

```{r}
# finds a percentage of benzene being inside the protein for each mutant
for (i in seq_along(main_df$name)) {
  directory <- paste(paste("data//bnz_distance/BNZ-COM-BNZ-AM-1-W15-110_129-", main_df$name[i], sep = ""), ".csv", sep = "")
  test_df <- read.csv(directory)
  names(test_df) <- c("Time", "Distance")
  test_df$isInside <- test_df$Distance<10
  main_df$isInsidePercent[i] <- length(test_df$isInside[test_df$isInside==TRUE])/5000
}
```



# Prediction model

```{r}
# Find the ones that are not in my original dataset
library("Peptides")
test_amino_acids <- setdiff(substr(list.files("data//d1_distance/"), 22, 23), main_df$name)
# Filter any non protein mutant combinations
test_amino_acids <- test_amino_acids[grepl("[ARDNCEQGHILKMFPSTWYV]{2}", test_amino_acids)]

for (aa in test_amino_acids) {
  df <- readXVG(paste(paste("data//d1_distance/D1-PRO-R1001-CG-2lcb-",aa,sep = ""),".xvg", sep = "")) 
  
  names(df) <- c("time", "distance")
  df$time <- as.numeric(df$time)
  df$distance <- as.numeric(df$distance)
      
      df$simulation <- case_when(df$time <= (100000 * 5) ~ 1,
                                 df$time <= (200000 * 5) ~ 2,
                                 df$time <= (299998 * 5) ~ 3)
      write.csv(
        x = df,
        file = paste("data//test_mutants/", paste(aa, ".csv", sep = ""), sep = ""),
        row.names = FALSE
      )
}
```

```{r}
# Save the main_df as csv file for future work
write.csv(
        x = main_df,
        file = paste("data//", paste("main_df", ".csv", sep = ""), sep = ""),
        row.names = FALSE
      )

```


```{r}
#Prediction 

# Build linear model 
model <- lm(mean_var ~ var_score, data = main_df)
# Add predictions 
pred.int <- predict(model, interval = "prediction")
mydata <- cbind(main_df, pred.int)
# Regression line + confidence intervals
p <- ggplot(mydata, aes(var_score, mean_var)) +
  geom_point() +
  stat_smooth(method = lm)
# Add prediction intervals
p + geom_line(aes(y = lwr), color = "red", linetype = "dashed")+
    geom_line(aes(y = upr), color = "red", linetype = "dashed") +
  labs(title = "Linear regression model plot") +
  scale_x_continuous(name = "Variance Score") +
  scale_y_continuous(name = "Mean Variance") +
  theme_minimal()

predict(model, newdata = data.frame(var_score = test_df$var_score))

test_df$predicted <- predict(model, newdata = data.frame(var_score = test_df$var_score))
```
```{r}
# The function that reads an individual simulation of a given mutant
get_sim <- function(AAname, simNum, data_directory = "data//extracted_data/") {
    fileName <- paste(AAname, ".csv", sep = "")
    path <- paste(data_directory, fileName, sep = "")
    
    df <- read.csv(path)
    
    if (simNum == 1) {
      return(df[c(10001:100000), ])
    } else if (simNum == 2) {
      return(df[c(110001:200000), ])
    } else if (simNum == 3) {
      return(df[c(209999:299998), ])
    }
}

# Function that gives all simulations concatenated
get_all_sim <- function(AAname, data_directory = "data//extracted_data/") {
    fileName <- paste(AAname, ".csv", sep = "")
    path <- paste(data_directory, fileName, sep = "")
    df <- read.csv(path)
    
    
    return(rbind(df[c(c(10001:100000), c(110001:200000), c(209999:299998)), ]))
}
```

```{r}
test_df <- data.frame(name = test_amino_acids)
test_df$first_letter <- substr(test_df$name,1,1)
test_df$second_letter <- substr(test_df$name,2,2)

sims_df_test <- rbind(data.frame(name = test_df$name, simulation = rep(1, length(test_df$name))),data.frame(name = test_df$name, simulation = rep(2, length(test_df$name))),data.frame(name = test_df$name, simulation = rep(3, length(test_df$name))))

for (i in seq_along(sims_df_test$name)) {
  sim_data <- get_sim(sims_df_test$name[i],sims_df_test$simulation[i], "data//test_mutants/")
  sims_df_test$variance[i] <- var(sim_data$distance)
}

sims_df_test <- sims_df_test[order(sims_df_test$name, sims_df_test$simulation),]

for (i in seq_along(test_df$name)) {
  aa <- test_df$name[i]
  df <- get_all_sim(aa, "data//test_mutants/")
  
  test_df$var_score[i] <-
      (aa_df[aa_df$amino_acid == test_df$first_letter[i],]$as_first_var +
      aa_df[aa_df$amino_acid == test_df$second_letter[i],]$as_second_var)
  
  
}
```



